Outline of Ideas for BTC Transaction Frequency

### 2015-10-07 outline/outline.tex
Added some description of the seasonal effects & trend components of the model

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

@@

To start, consider the following parameterized model using the structure discussed above. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\mu\delta_k,\beta)\\
\delta_{k-1} = \tau_k - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.

@@

For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. The key parameter we will control in these simulations is the mean transaction data rate, $\mu$, or the distribution of this rate. Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.

The simplest way to compute parameter estimates based on fitting model, e.g. maximum likelihood estimates, posterior means, or posterior medians. Then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\mu$, while holding other parameters constant.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\mu$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, suppose we fit the model and have access to the full (approximate) posterior distribution of the model parameters. Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\mu$'s distribution were different, e.g. its mean $\bar{\mu}$. We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\mu$ or $\bar{\mu}$ further away from the estimated values.

Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block $253,221$ from August 20, 2013 arriving at about 11:39 a.m.
% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct  5 14:56:54 2015
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
$\mu - \hat{\mu}$ & $P_{\bar{\mu}}$ & $P_{\mu^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
  \hline
0 & 0.01 & 0.01 & 0.00 \\ 
  0.01 & 0.03 & 0.03 & 0.03 \\ 
  0.05 & 0.23 & 0.23 & 0.22 \\ 
  0.1 & 0.42 & 0.42 & 0.42 \\ 
  0.5 & 0.82 & 0.82 & 0.83 \\ 
   \hline
\end{tabular}
\caption{Estimated probability of a filled block under a variety of assumptions. From the posterior distribution the mean is $\bar{\mu} = 0.0179$ and the median is $\mu^{(0.5)} = 0.0174$. The value of $\mu - \hat{\mu}$ is the amount added to the estimated value of $\mu$ for the purposes of the simulation, $P_{\bar{\mu}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\mu^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
\label{tab:probsims}
\end{table}

>>>>

While all L\'{e}vy processes are infinitely divisible, the gamma process is particularly convenient since the distribtion of $x_{\tau+\delta} - x_\tau$ and $x_\tau$ are both known in closed form as a consequence of the fact that if $x\sim G(\alpha_x,\beta)$ independent of $y\sim G(\alpha_y,\beta)$, then $x+y \sim G(\alpha_x + \alpha_y, \beta)$. Often L\'{e}vy processes are defined by considering an infinitely divisible distribution for $x_1$, but the distribution of $x_{\tau}$ for any $\tau>0$ need not be from the same class as $x_1$. The inverse Gaussian process is another L\'{e}vy process which has a closed form solution for the distribution of $x_\tau$. The inverse Gaussian density is
\[
p(x) = \left[\frac{\lambda}{2\pi x^3}\right]^{1/2}\exp\left[\frac{-\lambda(x - \mu)^2}{2\mu^2x}\right]
\]
for $x>0$ with mean parameter $\mu>0$, shape parameter $\lambda>0$, and variance $\mu^3/\lambda$. Then increments of the inverse Gaussian process are inverse Gaussian distributed, viz. $x_{\tau + \delta} - x_\tau \sim IG(\mu\delta,\lambda\delta^2)$.

@@

To start, consider the following parameterized model using the structure discussed above. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ --- the block has $\delta_k$ minutes to accumulate transaction data.

\subsection{Stochastic transaction rates --- seasonality and autocorrelation}
It is unlikely that the transaction rate is constant in time. The transaction rate is clearly growing as Bitcoin becomes more popular, and seasonal variation is expected throughout the data. Further, autocorrelation in transaction rates is likely since periods of high activity will sometimes be caused by, e.g., new information about the long term prospects of Bitcoin.

Let $\gamma_\tau$ denote the transaction rate at time $\tau$. We model the evolution of this rate as a variant of the exponential Ornstein–Uhlenbeck (OU) process. Specifically let $\varepsilon_t = \log(\gamma_t) - \mu_t$ , and let $\mu_t$ define a deterministic function of time containing seasonal and trend components. Then $\varepsilon_{\tau}$ is an OU process:
\begin{align*}
d\varepsilon_{\tau} = -\phi \varepsilon_{\tau}d\tau + \sigma_{\varepsilon} dB_{\tau}
\end{align*}
where $B_\tau$ is standard Brownian motion. It\^{o}'s lemma gives the stochastic differential equation for $\gamma_{\tau} = e^{\varepsilon_{\tau} + \mu_{\tau}}$:
\begin{align*}
d\gamma_\tau = -\phi(\log\gamma_\tau - \mu_\tau)\gamma_\tau + \sigma_{\varepsilon}^2/2\gamma_\tau +d\mu_\tau \gamma_\tau + \sigma_{\varepsilon}\gamma_\tau dB_\tau.
\end{align*}
The solution to the OU process is
\begin{align*}
\varepsilon_{\tau} = e^{-\phi\tau}\left(\varepsilon_0 + \int_0^\tau \sigma_{\varepsilon}e^{\phi s}dB_s\right)
\end{align*}
so that 
\begin{align*}
\varepsilon_{\tau + \delta} | \varepsilon_{\tau} \sim N\left(e^{-\phi\delta}\varepsilon_{\tau}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
\end{align*}
and thus
\begin{align*}
\log \gamma_{\tau + \delta}|\log\gamma_\tau \sim N\left(e^{-\phi\delta}(\log\gamma_{\tau} - \mu_\tau) + \mu_{\tau + \delta}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right).
\end{align*}
In other words, $\gamma_{\tau + \delta}$ has a lognormal distribution.

To stitch this piece onto the larger model, we exploit the inifnite divisibility property of the gamma distribution. Suppose that
\begin{align*}
x_{\tau + \delta} - x_\tau &\sim G(\beta \int_{\tau}^{\tau + \delta} d\gamma_s, \beta)
\end{align*}
with $d\gamma_\tau$ defined as above. Since both $x_{\tau}$ and $\gamma_{\tau}$ are continuous time processes, it should be possible to break the interval $[\tau, \tau + \delta]$ into a set of subintervals $[\tau + \delta_k, \tau + \delta_{k+1}]$, $0=\delta_0<\delta_1<\dots<\delta_K=\delta$ such that $x_{\tau + \delta} - x_{\tau} = \sum_{k=1}^K (x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}})$. Since $\int_{\tau}^{\tau + \delta}d\gamma_s = \gamma_{\tau + \delta} - \gamma_{\tau}$ we have 
\begin{align*}
x_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \stackrel{ind}{\sim} G(\beta(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta)
\end{align*}
so that by the gamma distribution's infinite divisibility property
\begin{align*}
\sum_{k=1}^Kx_{\tau + \delta_k} - x_{\tau + \delta_{k-1}} \sim G\left(\beta\sum_{k=1}^K(\gamma_{\tau + \delta_k} - \gamma_{\tau + \delta_{k-1}}), \beta\right) = G\left(\beta(\gamma_{\tau + \delta} - \gamma_\tau), \beta\right).
\end{align*}
In other words, this is consistent with the assumed distribution on $x_{\tau + \delta} - x_{\tau}$.

The full model can then be written as
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G\left(\beta(\gamma_{\tau_k + \delta_k} - \gamma_{\tau_k}),\beta\right)\\
\gamma_{\tau_k + \delta_k}|\gamma_{\tau_k} & \sim LN(\left(e^{-\phi\delta}(\log\gamma_{\tau_k} - \mu_{\tau_k}) + \mu_{\tau_k + \delta_k}, \frac{\sigma_{\varepsilon}^2}{2\phi}\left[1 - e^{-2\phi\delta}\right]\right)\\
\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
for blocks $k=1,2,\dots,N$ where $\mu_{\tau}$ is a parameterized deterministic function of $\tau$. For example, a trend with seasonal variability can be obtained via $\mu_{\tau} = \alpha_0 + \alpha_1\tau + \alpha_2\sin(\omega \tau)$ where $alpha_{0:2}$ is an uknown parameter and $\omega$ is the known frequency of the seasonal effect.

@@

For some distributional forms for the process $dx_\tau$ it may be possible to compute these quantities analytically, but in general it may be difficult. Using a Monte Carlo simulation of $N$ draws, we can approximate these probabilities by
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{\int_{\tau_k^{(i)}}^{\tau_{k}^{(i)}+\delta_k^{(i)}}dx_\tau^{(i)} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. The key parameter we will control in these simulations is the mean transaction data rate, $\gamma$, or the distribution of this rate. Starting with the basic model fit to, say, the last 1000 blocks on the blockchain (approximately one week of elapsed time), we can control this parameter in several ways.

The simplest way to compute parameter estimates based on fitting model, e.g. maximum likelihood estimates, posterior means, or posterior medians. Then estimate the probability of a block filling up by simulating the model conditional on these parameter values. In order to take into account increased Bitcoin usage, we can manually increase the mean transaction data rate parameter, $\gamma$, while holding other parameters constant.

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, suppose we fit the model and have access to the full (approximate) posterior distribution of the model parameters. Then we can simulate from the posterior of the parameters and for each simulated parameter, simulate the model in order to estimate the probability of a block filling up. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different, e.g. its mean $\bar{\gamma}$. We estimate the probability of a block filling using both approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.

Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block $253,221$ from August 20, 2013 arriving at about 11:39 a.m.
% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Mon Oct  5 14:56:54 2015
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
$\gamma - \hat{\gamma}$ & $P_{\bar{\gamma}}$ & $P_{\gamma^{(0.5)}}$ & $P_{\mathrm{post}}$ \\ 
  \hline
0 & 0.01 & 0.01 & 0.00 \\ 
  0.01 & 0.03 & 0.03 & 0.03 \\ 
  0.05 & 0.23 & 0.23 & 0.22 \\ 
  0.1 & 0.42 & 0.42 & 0.42 \\ 
  0.5 & 0.82 & 0.82 & 0.83 \\ 
   \hline
\end{tabular}
\caption{Estimated probability of a filled block under a variety of assumptions. From the posterior distribution the mean is $\bar{\gamma} = 0.0179$ and the median is $\gamma^{(0.5)} = 0.0174$. The value of $\gamma - \hat{\gamma}$ is the amount added to the estimated value of $\gamma$ for the purposes of the simulation, $P_{\bar{\gamma}}$ is the probability of a filled block using posterior means for parameter estimates, $P_{\gamma^{(0.5)}}$ is the probability of a filled block using posterior medians, and $P_{\mathrm{post}}$ is the probability of a filled block using the full posterior distribution.}
\label{tab:probsims}
\end{table}
